---
title: "EIT_Course_BigData"
author: "Gabriel Valenﾃｧa"
format: html
editor: visual
execute: 
  cache: false
---

## About the Data

Hourly pedestrian counts from automatic sensors in our local area.

Data is openly available at: <https://data.cityofsydney.nsw.gov.au/datasets/cityofsydney::automatic-hourly-pedestrian-count>

This dataset contains **hourly pedestrian** counts by location in Sydney, Australia.

**Automated counters** were placed in busy areas of the CBD to assess performance.

Data collection began in *February 2020* and ended in *June 2025*.

## Import Dataset

The dataset can be downloaded in many formats.

You can also import the data in many formats. Choose the one you prefer

-   CSV file

```{r}
data_csv <- read.csv("https://github.com/valenca13/EIT_Course_BigData/releases/download/Datasets/Automatic_Hourly_Pedestrian_Count.csv")
```

-   Shapefile

First Unzip file and place them into the ﾂｴDataﾂｴ file.

```{r, eval = FALSE}
#Unzip file

unzip("Data/Automatic_Hourly_Pedestrian_Count_Sydney.zip", junkpaths = TRUE)
```

```{r, eval= FALSE}
library(sf)
 
data_shapefile <- read_sf("Automatic_Hourly_Pedestrian_Count.shp")
```

## Convert dataset into *dataframe*

It is good practice to: \* convert the database into dataframe, as many packages only work with dataframes. \* To not work on the raw dataset.

```{r}
df <- data.frame(data_csv)
```

## Check the dataset

Structure of the dataset

```{r}
str(df)
```

Summary

```{r}
summary(df)
```
Convert the variable "Day" into a factor.

```{r}
df$Day <- factor(df$Day, labels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"), ordered = TRUE)

str(df$Day)
```

Check how many automatic counters there are:

```{r}
unique(df$Location_code)
```

```{r}
unique(df$Location_Name)
```

The location exact location of the automatic counters is not given in the dataset.

We need at least an approximation, to perform spatial analysis.

Check the description in the dashboard: https://experience.arcgis.com/experience/a0abf612051a4cc79918e6ce5ee7295b/

Go to Google Maps and retrieve the locations:

-   A001: -33.863518, 151.209968
-   A002: -33.871667, 151.210080
-   A003: -33.870783, 151.206607
-   A004: -33.873041, 151.207565

## Create variable with latitude and longitude of dataset

```{r}

df$Latitude[df$Location_code == "A001"] <- -33.863518
df$Longitude[df$Location_code == "A001"] <- 151.209968

df$Latitude[df$Location_code == "A002"] <- -33.871667
df$Longitude[df$Location_code == "A002"] <- 151.210080

df$Latitude[df$Location_code == "A003"] <- -33.870783
df$Longitude[df$Location_code == "A003"] <- 151.206607

df$Latitude[df$Location_code == "A004"] <- -33.873041
df$Longitude[df$Location_code == "A004"] <- 151.207565
```

## Create an *Geometry* variable

```{r}
library(sf)

df_sf <- st_as_sf(
  df,
  coords = c("Longitude", "Latitude"),
  crs = 4326
)
```

> **Note:** Google Maps uses *WGS84*, which corresponds to EPSG = 4326

Create an iterative map

```{r}
library(mapview)

mapview(df_sf, xcol="Latitude",ycol="Longitude")
```

## Have date and time in separate variables

Parse date-times with **y**ear, **m**onth, and **d**ay, **h**our, **m**inute, and **s**econd components.

```{r}
library(lubridate)

df_sf$Date <- ymd_hms(df_sf$Date, tz = "UTC")
```

Create new variables

```{r}
df_sf$Date_only <- as.Date(df_sf$Date)
df_sf$Year <- format(df_sf$Date, "%Y")
df_sf$Time_only <- format(df_sf$Date, "%H:%M:%S")
```

```{r}
class(df$Date)

class(df_sf$Date)
```

## Make some initial plots 

* Total pedestrian counts over time 

```{r}
library(ggplot2)

ggplot(df_sf, aes(x = Date, y = TotalCount)) +
  geom_line(linewidth = 2) +
    scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y"
  ) +
  labs(
    title = "Total Nﾂｺ of Pedestrians Over Time",
    x = "Time",
    y = "Total Nﾂｺ of Pedestrians"
  ) +
  theme_minimal()
```

* Total pedestrian counts per counter over time

```{r}
ggplot(df_sf, aes(x = Date, y = TotalCount, colour = Location_Name)) +
  geom_line() +
  facet_wrap(~ Location_code) + # insert "scales = "free_y" to adapt the maximum of Y in each plot
  theme_gray() +
  theme(axis.text.x = element_text(angle = 90, size = 6)) +
  scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y"
  ) +
  labs(
    title = "Total Nﾂｺ of Pedestrians by Automatic Counter Over Time",
    x = "Time",
    y = "Total Nﾂｺ of Pedestrians"
  )
```
_Facet_wrap_ divides the plots per automatic counter. 

* Tendency of total pedestrian counts per counter aggregated by year

```{r}
ggplot(df_sf, aes(x = Year, y = TotalCount, colour = Location_Name)) +
  geom_line(linewidth = 8) +
  facet_wrap(~ Location_code) + # insert "scales = "free_y" to adapt the maximum of Y in each plot
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, size = 7)) +
  scale_y_continuous(limits = c(0, max(df_sf$TotalCount) * 1.2))
```

Try changing the _themes_ for different appearances.  

> **Note:** Find here the documentation and ﾂｴcheatsheetﾂｴ of the ggplot2 package: https://ggplot2.tidyverse.org/


## Create a combined variables to analyze: 

* Day of Week + Year

```{r}
library(dplyr)



df_sf <- df_sf %>%
  mutate(
    Year = format(Date, "%Y"),
    DayYear = paste(Day, Year, sep = "_"),
    DayYear = factor(DayYear, levels = unique(DayYear))  # preserve order
  )
```

Plot

```{r}
ggplot(df_sf, aes(x = DayYear, y = TotalCount, colour = Day)) +
  geom_line(size =3) +
  # geom_point(size = 2) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, size = 7)
  )
```






Criar uma variavel de juntar por dia semana e por ano.

mutate e um groupby e summarize para fazer a soma de total counts.

grafico para os 5 dias da semana.



## Import land use data and integrate with existing data

https://data.cityofsydney.nsw.gov.au/pages/open-data#Categories

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
