---
title: "Data treatment"
format: html
editor: visual
execute: 
  cache: false
---

## About the Data

Hourly pedestrian counts from automatic sensors in our local area.

Data is openly available at: <https://data.cityofsydney.nsw.gov.au/datasets/cityofsydney::automatic-hourly-pedestrian-count>

This dataset contains **hourly pedestrian** counts by location in Sydney, Australia.

**Automated counters** were placed in busy areas of the CBD to assess performance.

Data collection began in *February 2020* and ended in *June 2025*.

## Import Dataset

The dataset can be downloaded in many formats.

You can also import the data in many formats. Choose the one you prefer

-   CSV file

```{r}
data_csv <- read.csv("https://github.com/valenca13/EIT_Course_BigData/releases/download/Datasets/Automatic_Hourly_Pedestrian_Count.csv")
```

-   Shapefile

First Unzip file and place them into the ´Data´ file.

```{r, eval = FALSE}
#Unzip file

unzip("Data/Automatic_Hourly_Pedestrian_Count_Sydney.zip", junkpaths = TRUE)
```

```{r, eval= FALSE}
library(sf)
 
data_shapefile <- read_sf("Automatic_Hourly_Pedestrian_Count.shp")
```

## Convert dataset into *dataframe*

It is good practice to: \* convert the database into dataframe, as many packages only work with dataframes. \* To not work on the raw dataset.

```{r}
#| warning: false
library(tidyverse)
df <- data.frame(data_csv)
```

## Check the dataset

* Structure of the dataset

```{r}
str(df)
```

* Summary of the dataset

```{r}
library(skimr)
skim(df)
```

## Check Missing data

```{r}
table(is.na(df))
```
* Plot missing data

```{r}
library(DataExplorer)

plot_missing(df, plotly = TRUE)
```
* **Listwise deletion**. Removes an entire row (case) if any variable in that row is missing. Only fully complete observations remain.
    
```{r}
df_missingListwise = na.omit(df) #removes all rows with at least one NA in any variable
```

```{r}
plot_missing(df_missingListwise)
```

* **Pairwise deletion**. Delete the row of missing value to the presence in a the specific variable involved in a given calculation.
    
```{r}
df_missingPairwise = df[!is.na(df$Previous52DayTimeAvg),] #removes all rows with NA in Previous52DayTimeAvg variable
```

For our case, which treatment would be better? 

> **Note:** There are many methods for treating missing data, such as replacing the values by the mean, median or using more complex imputation methods (e.g. k‑Nearest Neighbors (kNN) Imputation). 

## Treat existent variables

* Convert the variable "Day" into a factor.

```{r}
df$Day <- factor(df$Day, labels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"), ordered = TRUE)

str(df$Day)
```
Check how many automatic counters there are:

```{r}
unique(df$Location_code)
```

```{r}
unique(df$Location_Name)
```

The location exact location of the automatic counters is not given in the dataset.

We need at least an approximation, to perform spatial analysis.

Check the description in the dashboard: <https://experience.arcgis.com/experience/a0abf612051a4cc79918e6ce5ee7295b/>

Go to Google Maps and retrieve the locations:

-   A001: -33.863518, 151.209968
-   A002: -33.871667, 151.210080
-   A003: -33.870783, 151.206607
-   A004: -33.873041, 151.207565

## Create variable with latitude and longitude of dataset

```{r}

df$Latitude[df$Location_code == "A001"] <- -33.863518
df$Longitude[df$Location_code == "A001"] <- 151.209968

df$Latitude[df$Location_code == "A002"] <- -33.871667
df$Longitude[df$Location_code == "A002"] <- 151.210080

df$Latitude[df$Location_code == "A003"] <- -33.870783
df$Longitude[df$Location_code == "A003"] <- 151.206607

df$Latitude[df$Location_code == "A004"] <- -33.873041
df$Longitude[df$Location_code == "A004"] <- 151.207565
```

## Create an *Geometry* variable

```{r}
library(sf)

df_sf <- st_as_sf(
  df,
  coords = c("Longitude", "Latitude"),
  crs = 4326
)
```

> **Note:** Google Maps uses *WGS84*, which corresponds to EPSG = 4326

Create an iterative map

```{r}
library(mapview)

df_map <- unique(df_sf$geometry)
mapview(df_map)
```

## Have date and time in separate variables

Parse date-times with **y**ear, **m**onth, and **d**ay, **h**our, **m**inute, and **s**econd components.

```{r}
#| warning: false
library(lubridate)

df_sf$Date <- ymd_hms(df_sf$Date, tz = "UTC")
```

Check the class of the variable ´Date´

```{r}

class(df$Date)

class(df_sf$Date)
```

Create new variables separating Date, Year, and Time

```{r}
df_sf$Date_only <- as.Date(df_sf$Date)
df_sf$Year <- format(df_sf$Date, "%Y")
df_sf$Time_only <- format(df_sf$Date, "%H:%M:%S")
```

## Make some initial plots

* Total pedestrian counts over time

```{r}
library(ggplot2)

ggplot(df_sf, aes(x = Date, y = TotalCount)) +
  geom_line(linewidth = 1) +
    scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y"
  ) +
  labs(
    title = "Total Nº of Pedestrians Over Time",
    x = "Time",
    y = "Total Nº of Pedestrians"
  ) +
  theme_minimal()
```

* Total pedestrian counts per counter over time

```{r}
ggplot(df_sf, aes(x = Date, y = TotalCount, colour = Location_Name)) +
  geom_line() +
  facet_wrap(~ Location_code) + # insert "scales = "free_y" to adapt the maximum of Y in each plot
  theme_gray() +
  theme(axis.text.x = element_text(angle = 90, size = 6)) +
  scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y"
  ) +
  labs(
    title = "Total Nº of Pedestrians by Automatic Counter Over Time",
    x = "Time",
    y = "Total Nº of Pedestrians"
  )
```

*Facet_wrap* divides the plots per automatic counter.

* Tendency of total pedestrian counts per counter aggregated by year

```{r}
ggplot(df_sf, aes(x = Year, y = TotalCount, colour = Location_Name)) +
  geom_line(linewidth = 4) +
  facet_wrap(~ Location_code) + # insert "scales = "free_y" to adapt the maximum of Y in each plot
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, size = 7)) +
  scale_y_continuous(limits = c(0, max(df_sf$TotalCount) * 1.2))
```
* Total number of pedestrians per time

```{r}
ggplot(df_sf, aes(x = Time_only, y = TotalCount, colour = Location_Name)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ Location_code, scales ="free_y") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, size = 5, hjust = 1)) +
  scale_y_continuous(limits = c(0, max(df_sf$TotalCount) * 1.2))
```

Try changing the *themes* for different appearances.

> **Note:** Find here the documentation and ´cheatsheet´ of the ggplot2 package: https://ggplot2.tidyverse.org/

## Create a combined variables to analyze:

-   Day of Week + Year

```{r}
library(dplyr)



df_sf <- df_sf %>%
  mutate(
    Year = format(Date, "%Y"),
    DayYear = paste(Day, Year, sep = "_"),
    DayYear = factor(DayYear, levels = unique(DayYear))  # preserve order
  )
```

Plot

```{r}
ggplot(df_sf, aes(x = DayYear, y = TotalCount, colour = Day)) +
  geom_line(size =3) +
  # geom_point(size = 2) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, size = 7)
  )
```

Criar uma variavel de juntar por dia semana e por ano.

mutate e um groupby e summarize para fazer a soma de total counts.

grafico para os 5 dias da semana.

## Import land use data and integrate with existing data

https://data.cityofsydney.nsw.gov.au/pages/open-data#Categories

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
