---
eval: true
code-fold: false
number-depth: 2
engine: knitr
---

# R basics

In this chapter we will introduce to the R basics and some exercises to get familiar to how R works.

## Math operations

**Sum**

```{r}
1+1
```

**Subtraction**

```{r}
5-2
```

**Multiplication**

```{r}
2*2
```

**Division**

```{r}
8/2
```

**Round the number**

```{r}
round(3.14)
round(3.14, 1) # The "1" indicates to round it up to 1 decimal digit.
```

You can use help `?round` in the console to see the description of the function, and the default arguments.

## Basic shortpaths

**Perform Combinations**

```{r}
c(1, 2, 3)
c(1:3) # The ":" indicates a range between the first and second numbers. 
```

**Create a comment with `ctrl + shift + c` **

```{r}
# Comments help you organize your code. The software will not run the comment. 
```

## Create a table

A simple table with the number of trips by car, PT, walking, and cycling in a hypothetical street segment at a certain period.

**Define variables**

```{r}
# Transport mode choice
modes <- c("car", "PT", "walking", "cycling") # you can use "=" or "<-"

# Number of trips
Trips = c(200, 50, 300, 150) # Be careful with uppercase letters. 

# Trip Origin
origin <- c("Home", "Home", "School", "Grocery")

# Trip Destination
destination <- c("Work", "University", "Park", "Home")
```

**Join the variables to create a table**

```{r}
table_example = data.frame(modes, Trips, origin, destination)
```

**Take a look at the table**

Visualize the table by clicking on the "table_example" in the "Environment" page or use:

```{r}
View(table_example)
```

**Look at the first row**

```{r}
table_example[1,] #rows and columns start from 1 in R, differently from Python which starts from 0.
```

**Look at first row and column**

```{r}
table_example[1,1]
```

::: {.callout-tip appearance="simple"}
After you type `"` you can use `tab` to navigate between folders and files and `enter` to autocomplete.
:::

## Explore the data

**Check the total number of trips**

Use `$` to select a variable of the data

```{r}
sum(table_example$Trips)
```
**Filter by condition (create new tables)**

Filter trips only with origin from Home

```{r}
data_Lisbon = table_example[table_example$origin == "Home",]
```

Filter trips with origin **different** from Home

```{r}
data_out_Lisbon = table_example[table_example$origin != "Home",]
```

**Remove the first column**

```{r}
data = data[ ,-1] #first column
```

**Create a table only with origin, destination**

There are many ways to do the same operation.

```{r}
names(data)
```

```{r}
data_OD = table_example[ ,c(3,4)]
```

```{r}
data_OD2 = table_example[ ,-c(1:2)] #Or use a comma. 
```

```{r}
#| include: false
#| eval: false
# this coverts this quarto to a plain r script
knitr::purl("r-basics.qmd", "code/r-basics.R", documentation = 2)

```
