---
title: "Spatial Data Analysis"
---
## Import libraries

```{r}
#| warning: false
# library(osmdata)
# library(sf)
library(dplyr)
# library(terra)
library(ggplot2)
library(tidyverse)
library(osmextract)
```

We will work with **OpenStreetMap**. Check the available **features** and **tags** through the following [link](https://wiki.openstreetmap.org/wiki/Map_features#Highway ). 

## Import OpenStreetMap data

```{r}
OSM_Sydney = oe_get_network(place = "Sydney") # it will geocode the place
```

* Plot the entire network

```{r}
par(mar = rep(0.1, 4))
plot(sf::st_geometry(OSM_Sydney))
```

* Check for missing data in 'Highway'

```{r}
table(is.na(OSM_Sydney$highway))
```

* Check for the values in 'Highway'

```{r}
unique(OSM_Sydney$highway)
```

## Clean the dataset

* **Filter the dataset only for the main 'Highway' classifications** 

```{r}
Sydney_main_roads = OSM_Sydney %>% 
  filter(highway %in% c("motorway", "trunk", "primary", "secondary", "tertiary", "residential"))
```

> **Note:** In order to check what are the main categories in OSM, check the [link](https://wiki.openstreetmap.org/wiki/Key:highway#Highway

**Remove variables that are not useful** 

Check the order of the variables

```{r}
names(OSM_Sydney)
```
```{r}
Sydney_main_roads <- Sydney_main_roads[,-c(4:12)]
```

* **Plot the filtered network**

```{r}
par(mar = rep(0.1, 4))
plot(sf::st_geometry(Sydney_main_roads))
```
## Centrality measures

* Betweenness centrality








* **Import Land-use Data** 

