---
title: "Spatial Data Analysis"
---

In this chapter we will learn how to extract and merge _OpenStreetMap_ data of amenities and road network to the pedestrian count data. 


## Import libraries

```{r, eval = FALSE}
#| warning: false
library(osmdata)
library(sf)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(osmextract)
```

## Extract land use amenities

* **Define Location**

```{r}
location <- "Sydney, Australia"
```

* **Define the municipality boundary**

```{r}
# sydney_boundary <- opq(location) |>
#   add_osm_feature(key = "admin_level", value = "8") |>   # municipalities
#   osmdata_sf()
```

admin_level = 8 corresponds to local government areas (municipalities) in Australia.

* **Extract the poligon**

```{r}
# # Extract the polygon
# sydney_poly <- sydney_boundary$osm_polygons |> 
#   st_union() |> 
#   st_make_valid()
```

* Create an OSM query for extracting amenities data

```{r}
amenities_query <- opq(location) |> 
  add_osm_feature(key = "amenity", value = c("restaurant", "supermarket", "bar")) 
```

> **Note:** We will use only two categories for simplicity of the example. Ideally, more types of amenities should be considered for analysis. 

* Extract the data

```{r}
# Set an alternative Overpass server
set_overpass_url("https://lambert.openstreetmap.de/api/interpreter")

amenities_data <- osmdata_sf(amenities_query)
```

* View the first few rows of the amenities data

```{r}
head(amenities_data$osm_points)
```

* Treat dataset

```{r}
names(amenities_data$osm_points)
```

```{r}
amenities_data_treated <- amenities_data$osm_points[,c(1,2,26,228)]
```

* Explore dataset

```{r}
unique(amenities_data_treated$amenity)
```

* **Save dataset** 

```{r}
# st_write(amenities_data_treated, "amenities.shp", delete_dsn = TRUE)
```

We will work with **OpenStreetMap**. Check the available **features** and **tags** through the following [link](https://wiki.openstreetmap.org/wiki/Map_features#Highway).

## Import OpenStreetMap data

```{r}
OSM_Sydney = oe_get_network(place = "Sydney") # it will geocode the place
```

-   Plot the entire network

```{r}
par(mar = rep(0.1, 4))
plot(sf::st_geometry(OSM_Sydney))
```

-   Check for missing data in 'Highway'

```{r}
table(is.na(OSM_Sydney$highway))
```

-   Check for the values in 'Highway'

```{r}
unique(OSM_Sydney$highway)
```

## Clean the dataset

-   **Filter the dataset only for the main 'Highway' classifications**

```{r}
Sydney_main_roads = OSM_Sydney %>% 
  filter(highway %in% c("motorway", "motorway_link", "primary", "primary_link", "secondary","secondary_link", "trunk", "trunk_link")) 
```

Try adding the "tertiary", "residential" streets, and the network gets more complex

> **Note:** In order to check what are the main categories in OSM, check the [link](https://wiki.openstreetmap.org/wiki/Key:highway#Highway)

**Remove variables that are not useful**

Check the order of the variables

```{r}
names(OSM_Sydney)
```

```{r}
Sydney_main_roads <- Sydney_main_roads[,-c(4:12)]
```

-   **Plot the filtered network**

```{r}
par(mar = rep(0.1, 4))
plot(sf::st_geometry(Sydney_main_roads))
```

* **Plot the amenities with the network**

```{r}
ggplot() +
  geom_sf(data = Sydney_main_roads, color = "black", size = 0.4) +
  geom_sf(data = amenities_data_treated, aes(color = amenity), size = 1) +
  theme_minimal() +
  labs(title = "Amenities in Sydney", color = "Amenity Type") +
  theme(legend.position = "bottom")
```
## Import pedestrian counting data (assuming it's in a CSV file with geometry)

```{r}
pedestrian_data <- read.csv("pedestrian_1.csv")
```

## Create a grid over the area of interest



* # Count the number of pedestrians in each grid cell








## Centrality measures

-   Betweenness centrality

```{r}
library(igraph)
library(sf)
library(dplyr)
```

Give each edge an unique index

```{r}
edges <- Sydney_main_roads %>%
  mutate(edgeID = c(1:n()))
```

Create nodes at the start and end point of each edge

```{r}
nodes <- edges %>%
  st_coordinates() %>%
  as_tibble() %>%
  rename(edgeID = L1) %>%
  group_by(edgeID) %>%
  slice(c(1, n())) %>%
  ungroup() %>%
  mutate(start_end = rep(c('start', 'end'), times = n()/2))

head(nodes) #show the first values
```

Give each node an unique index

```{r}
nodes <- nodes |>
  mutate(
    xy = paste(X, Y),
    nodeID = as.integer(factor(xy))
  ) |>
  select(-xy)

head(nodes)
```

Combine the node indices with the edges

```{r}
source_nodes <- nodes %>%
  filter(start_end == 'start') %>%
  pull(nodeID)

target_nodes <- nodes %>%
  filter(start_end == 'end') %>%
  pull(nodeID)

edges = edges %>%
  mutate(from = source_nodes, to = target_nodes)

edges
```

Remove duplicate nodes

```{r}
nodes <- nodes %>%
  distinct(nodeID, .keep_all = TRUE) %>%
  select(-c(edgeID, start_end)) %>%
  st_as_sf(coords = c('X', 'Y')) %>%
  st_set_crs(st_crs(edges))

nodes
```

Convert to tbl_graph

```{r}
library(tidygraph)
```

```{r}
graph = tbl_graph(nodes = nodes, edges = as_tibble(edges), directed = FALSE)

graph
```

Putting it together

```{r}


# sf_to_tidygraph(Sydney_main_roads, directed = FALSE)
```

Plot with activate funciton

```{r}
library(tmap)
```

```{r}
tmap_mode('view')

tm_shape(graph %>% activate(edges) %>% as_tibble() %>% st_as_sf()) +
  tm_lines() +
tm_shape(graph %>% activate(nodes) %>% as_tibble() %>% st_as_sf()) +
  tm_dots() +
tmap_options(basemaps = 'OpenStreetMap')
```

Centrality measures

```{r}
library(tidygraph)

graph <- as_tbl_graph(graph)
```

Centrality

```{r}
graph <- graph %>%
  activate(edges) %>%
  mutate(length = st_length(geometry))
```

```{r}
graph <- graph %>%
  activate(nodes) %>%
  mutate(degree = centrality_degree()) %>%
  mutate(betweenness = centrality_betweenness(weights = length)) %>%
  activate(edges) %>%
  mutate(betweenness = centrality_edge_betweenness(weights = length))

```

```{r}
ggplot() +
  geom_sf(data = graph %>% activate(edges) %>% as_tibble() %>% st_as_sf(), col = 'grey50') + 
  geom_sf(data = graph %>% activate(nodes) %>% as_tibble() %>% st_as_sf(), aes(col = betweenness, size = betweenness)) +
  scale_colour_viridis_c(option = 'inferno') +
  scale_size_continuous(range = c(0,4))
```

The geometry is a "Linestring", convert it to edges

```{r}
sf_lines <- st_as_sf(Sydney_main_roads, wkt = "geometry")
```

```{r}
edges <- sf_lines |>
  st_coordinates() |>
  as.data.frame() |>
  group_by(L1) |>
  mutate(next_x = lead(X),
         next_y = lead(Y)) |>
  filter(!is.na(next_x)) |>
  transmute(
    from = paste(X, Y, sep = ","),
    to   = paste(next_x, next_y, sep = ",")
  )
```

Build igraph object

```{r}
g <- graph_from_data_frame(edges, directed = FALSE)
```

Centrality measures

```{r}
degree <- degree(g)
```

```{r}
closeness <- closeness(g)
```

```{r}
betweenness <- betweenness(g)
```

```{r}
centrality <- data.frame(
  node = names(degree),
  degree = degree,
  betweenness = betweenness,
  closeness = closeness
)
```

```{r}
plot(
  g,
  vertex.size = scales::rescale(degree, to = c(8, 30)),
  vertex.label.cex = 0.8,
  main = "Graph with Node Size = Degree"
)
```

-   **Import Land-use Data**
