---
title: "Spatial Data Analysis"
---

In this chapter we will learn how to extract and merge _OpenStreetMap_ data of amenities and road network to the pedestrian count data. 


## Import libraries

```{r, eval = FALSE}
#| warning: false
library(osmdata)
library(sf)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(osmextract)
```

## Extract land use amenities

* **Define Location**

```{r}
location <- "Sydney, Australia"
```

* Create an OSM query for extracting amenities data

```{r}
library(osmdata)
amenities_query <- opq(location) |> 
  add_osm_feature(key = "amenity", value = c("restaurant", "bank", "bar")) 
```

> **Note:** We will use only two categories for simplicity of the example. Ideally, more types of amenities should be considered for analysis. 

* Extract the data

```{r}
# Set an alternative Overpass server
set_overpass_url("https://overpass-api.de/api/interpreter")

amenities_data <- osmdata_sf(amenities_query)
```

* View the first few rows of the amenities data

```{r}
head(amenities_data$osm_points)
```

* Treat dataset

```{r}
names(amenities_data$osm_points)
```

```{r}
amenities_data_treated <- amenities_data$osm_points[,c(1,2,26,228)]
```

* Explore dataset

```{r}
unique(amenities_data_treated$amenity)
```



We will work with **OpenStreetMap**. Check the available **features** and **tags** through the following [link](https://wiki.openstreetmap.org/wiki/Map_features#Highway).

## Import OpenStreetMap data

```{r}
library(osmextract)
OSM_Sydney = oe_get_network(place = "Sydney") # it will geocode the place
```

-   Plot the entire network

```{r}
par(mar = rep(0.1, 4))
plot(sf::st_geometry(OSM_Sydney))
```

-   Check for missing data in 'Highway'

```{r}
table(is.na(OSM_Sydney$highway))
```

-   Check for the values in 'Highway'

```{r}
unique(OSM_Sydney$highway)
```

## Clean the dataset

-   **Filter the dataset only for the main 'Highway' classifications**

```{r}
Sydney_main_roads = OSM_Sydney |> 
  dplyr::filter(highway %in% c("motorway", "motorway_link", "primary", "primary_link", "secondary","secondary_link", "trunk", "trunk_link")) 
```

Try adding the "tertiary", "residential" streets, and the network gets more complex

> **Note:** In order to check what are the main categories in OSM, check the [link](https://wiki.openstreetmap.org/wiki/Key:highway#Highway)

**Remove variables that are not useful**

Check the order of the variables

```{r}
names(OSM_Sydney)
```

```{r}
Sydney_main_roads <- Sydney_main_roads[,-c(4:12)]
```

-   **Plot the filtered network**

```{r}
par(mar = rep(0.1, 4))
plot(sf::st_geometry(Sydney_main_roads))
```

* **Plot the amenities with the network**

```{r}
library(ggplot2)
ggplot() +
  geom_sf(data = Sydney_main_roads, color = "black", size = 0.4) +
  geom_sf(data = amenities_data_treated, aes(color = amenity), size = 1) +
  theme_minimal() +
  labs(title = "Amenities in Sydney", color = "Amenity Type") +
  theme(legend.position = "bottom")
```
## Import pedestrian counting data (assuming it's in a CSV file with geometry)

```{r}
pedestrian_data <- read.csv("pedestrian_1.csv")
```

## Create a grid over the area of interest


